Predicting how climate change will affect terrestrial mammals' movement and habitat selection

### 2024-06-13 writing/manuscript.Rmd
edited intro

---
title: "Predicting how climate change and short-term weather will affect terrestrial mammals' movement and habitat selection"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Kirk Safford
    institute: bcp
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammal-speeds.bib'
# csl: 'csl.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

\doublespacing
*
```{r setup, include=FALSE}
# set chunk defaults: don't print code, images text width and centered
knitr::opts_chunk$set(echo = FALSE, out.width = '\\textwidth',
                      fig.align = 'center', cache = TRUE)
```

@@

@@

@@

Intro:

* add references:
  * McNabb 1960s
  * Hirt 2022
  * Moose speed and temp (just @alston_temperature_2020 or are there others?)
  * Lots of work on fish
  * Brown temp constant for all biological processes
  * Andean hummingbird and temp
* address any questions answered in the discussion

Methods:

* add headers

Write abstract

@@

The rate at which chemical reactions occur within an organism [metabolism, see @brown_toward_2004] depends strongly on the temperature reactions occur at. However, organisms experience different ranges of external temperatures and, consequently, exhibit specific thermal optima. These depend on many factors, including body size [@taylor_energetics_1982; @reiss_scaling_1988; @fjelldal_nightly_2021], physiological adaptations [@schmidt_interplay_2020; @mohr_cellular_2020], and behavior [@arechavala-lopez_common_2019]. In the case of heterothermic organisms, low temperatures increase energetic costs [@melin_effect_2023; @ref; @fish; @andean_hummingbird], while high temperatures increase the risk of overheating [@ref; @cheetas; @alston_temperature_2020].

At extremely cold temperatures, the metabolic costs of movement and homeothermy increase energetic expenditures, which can constrain an animal's ability to move, lowering encounter rates with food [@hou_cold_2020], fitness [@ref], and, ultimately, chance of survival [@berger_climate_2018]. Mammals reduce the adverse effects of cold through a combination of both behavioral adaptations [e.g., searching for shelter: @ref; reducing movement rates: @ref] and physiological adaptations [e.g., fur: @ref; circulation: @ref]. At extremely hot temperatures, excessively high body temperatures ***XXXX*** [@alston_temperature_2020; @refs].

Evolutional constraints limit mammals' ability to respond temperature extremes, which results in phylogenetically-constrained temperature optima for different species and populations [@caribou-ref]. Thus, while animals are adapted to fluctuations around average conditions [@ref], strong and consistent temperature anomalies can decrease an animal's energetic stability [@ref], reproductive success [@schmidt_interplay_2020], and ability to move [@ref]. Recent changes in climate have resulted in widespread warming, milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ref_2021_heat_dome]. These changes are expected to affect mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress [@weststrate_how_2024].

This paper provides a BC-focused analysis on the effects of proximate weather and long-term climate on mammal movement. Using more than a decade of telemetry data and proximal 3-hour weather data, we estimate how animals altered their movement frequency, movement speed, and habitat selection in response to air temperature. We then pair these responses with climate change projections to forecast changes mammals movement in the current century under different climate change scenarios [@riahi_shared_2017]. The analyses include data from six species (*Canis lupus*, *Cervus elaphus*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, *Ursus arctos horribilis*; see Table \@ref(tab:data-table)) across a large spatial range of British Columbia (Fig. \@ref(fig:bc-map)). We discuss the consequences of changes in mammalian movement in relationship to energetics, food encounter rates, and the incidences of human-wildlife.

@@

```{r data-table}
readRDS('../models/movement-models-2024-03-18.rds') %>%
  transmute(Dataset = if_else(grepl('Puma', dataset_name),
                                'Puma concolor', dataset_name) %>%
              stringr::str_replace_all('_', ' ') %>%
              stringr::str_replace_all(' boreal', ' (boreal)') %>%
              stringr::str_replace_all(' southern mountain',
                                       ' (southern mountain)'),
            animal,
            tel = purrr::map(tel, data.frame)) %>%
  unnest(tel) %>%
  group_by(animal) %>%
  mutate(dt = timestamp - lag(timestamp)) %>%
  group_by(Dataset) %>%
  summarize(Start = min(as.Date(timestamp)),
            End = max(as.Date(timestamp)),
            dt = as.numeric(median(dt, na.rm = TRUE))) %>%
  mutate(dt = round('hours' %#% dt, 2)) %>%
  rename(`Median sampling interval (hours)` = dt) %>%
  knitr::kable(format = 'pipe', caption = 'Start and end of the GPS telemetries along with the median sampling interval for each of the six species included in this study.') %>%
  kableExtra::column_spec(1, italic = TRUE) %>%
  kableExtra::kable_styling(font_size = 10)
```

@@

Animal telemetry datasets were obtained from multiple sources. Elk tracking data were obtained from Movebank [@kays_movebank_2022; study name: Elk in southwestern Alberta, see @study_ref], while the boreal caribou and wolf data were obtained from a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/)). ***All other tracking data were obtained from private collaborators*** [@refs_to_studies]. Outlier GPS locations were removed following diagnostic analyses of the location of each point as well as straight-line displacement and turning angle between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$), especially if points before or after the potential outlier indicated stationary behavior. The script used to clean the data and all custom functions used are available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-speeds](https://github.com/QuantitativeEcologyLab/bc-mammals-speeds)).

@@

Rasters of three-hour reanalyzed air temperature data were downloaded from the ERA5 dataset [@copernicus_climate_change_service_era5_2019] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)). Proximal weather conditions were then estimated for each location by extracting the values from the corresponding cell from the temporally nearest raster using the `extract()` function from the `terra` package for `R` [version 1.7-46, @hijmans_terra_2023].

@@

Since `ctmm`'s movement models allow for stochastic speeds but assume constant movement, speeds were set to zero if they were not significantly different from zero ($\alpha = 0.05$), under the assumption of a $\chi$-distributed speeds. ***NOT SURE IF THIS IS CORRECT*** The function used is available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-speeds/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-speeds/blob/main/functions/detrend_speeds.R)). To avoid artifacts due to excessively small, non-zero speeds, estimated speeds were then clustered into two groups using a $k$-means algorithm with two centers (see Appendix A).

@@

We estimated the effects of hourly temperature on mammals' movement using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_pedersen_hierarchical_2019] via `mgcv` package for `R`. The first HGAM had a binomial family of distribution and estimated the probability of a mammal moving as a function of time of day, day of year, and temperature. Since there was substantial variation in the direction and wiggliness of each species' trends, the effects were modeled independently `by` smooths for each species [see model I in Figure 4 of @pedersen_hierarchical_2019]. The model accounted for the cyclicity of time of day and day of year using cyclic cubic splines, which provide continuity in the smooth term by ensuring the function and its slope (first derivative) and curvature (second derivative) are continuous between consecutive cycles. The second HGAM had the same structure, but it used a Gamma family of distributions to estimate the speed of an animal given that it was moving. Together, the two HGAMs inform us on an animal's long-term speed (i.e., distance traveled per unit time), since it is the product of probability that the animal is moving and the speed it moves at, given that it is moving.

@@

An animal's habitat selection is proportional to the number of times one can observe it in an environment. Consequently, one can view the spatial distributions of an animal in space as a poisson process whose mean depends on the amount of space, $\Delta u$, and time, $\Delta t$, the observation takes place over, which we can write as $\lambda(\Delta u, \Delta t)$. If we shrink both $\Delta u$ and $\Delta t$ to a single point in space and a single moment in time, we now have $\lambda(u, t) \le 1$, since we cannot observe more than one animal in a given point in time and spce $(u, t)$. Additional theoretical background is provided by @aarts_estimating_2008 and @alson_mitingting_2022 and in Appendix B.

@@

* energetics
* encounter rates
* HWI

>>>>

---
title: "Predicting how climate change will affect terrestrial mammals' movement and habitat selection"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Kirk Safford
    institute: bcp
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
# csl: 'csl.csl'
# csl: 'nature.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

\doublespacing

```{r setup, include=FALSE}
# set chunk defaults: don't print code, images text width and centered
knitr::opts_chunk$set(echo = FALSE, out.width = '\\textwidth',
                      fig.align = 'center', cache = TRUE, warning = FALSE,
                      message = FALSE)
```

@@

```{r, echo=FALSE}
abstract <- ""
```

<!-- elements of the title page -->

\noindent \textbf{Article type}: Major article

\noindent \textbf{Words in abstract}: `r stringi::stri_count_words(abstract)`

\noindent \textbf{Words in main text}: `r suppressMessages(wordcountaddin::word_count())`

\noindent \textbf{Figures}: 2

\noindent \textbf{Tables}: 1

\noindent \textbf{References}: 33 (updated on 2024-06-13) <!-- counted by using nature csl -->

\noindent \textbf{Appendices}: 3

\noindent \textbf{Key words:} energetics, environmental stochasticity, home range, range size, `ctmm`, simulations

<!-- Authors' emails:
Stefano Mezzini: stefano.mezzini@ubc.ca
Chris H. Fleming: christen.fleming@ucf.edu
E. Patrícia Medici: medici@ipe.org.br
Michael J. Noonan: michael.noonan@ubc.ca
-->

@@

\noindent `r abstract`

\newpage

@@

@@

Animals' energetic costs depend strongly on ambient temperature [@brown_toward_2004]. Cold temperatures increase the metabolic costs of movement [@taylor_energetics_1982] and homeothermy [@melin_effect_2023], which can constrain an animal's ability to move [@schwerdt_ecological_2024], its fitness [@walker_global_2019], and, ultimately, its odds of survival [@berger_climate_2018]. Hot temperatures tend to lower energetic needs [@taylor_energetics_1982], but they also cause animals to overheat, which lowers their ability to move [@powers_hovering_2017; @alston_temperature_2020; @dyer_travel_2023].

Mammals reduce the adverse effects of extreme temperatures through both physiological and behavioral adaptations [@jessen_temperature_2001; @mota-rojas_physiological_2021], but their ability to respond through physiological adaptations is limited by their evolutionary history, since their optimal temperature and range of tolerable temperatures (Fig. \@ref(fig:temperature-optima)) depend on the environment they live in [@levins_evolution_1974]. In contrast, changes in movement rates and space use allow mammals to respond to stressors immediately [@alston_temperature_2020; @dyer_travel_2023; @attias_effects_2018; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].

```{r temperature-optima, fig.width=6, fig.height=4, out.width='50%', fig.cap="Hypothetical ranges in fitness as a consequence of ambient temperature for a cold-adapted mammal (blue), temperate-adapted mammal (yellow), and hot-adapted mammal (red). Contrast the differences in fitness gains at different temperatures, as well as differences in the height and spead of the three curves."}
library('dplyr')
library('tidyr')
library('ggplot2')
library('khroma')
theme_set(theme_bw() + theme(panel.grid = element_blank()))

tibble(x = seq(-30, 30, by = 0.01),
       cold = rev(dgamma(x + 12, shape = 10, rate = 0.5)),
       mid = dnorm(x, mean = 5, sd = 5),
       warm = dgamma(x, shape = 56, rate = 2.8)) %>%
  pivot_longer(-c(x)) %>%
  ggplot() +
  geom_area(aes(x, value, color = name, fill = name), alpha = 0.2,
            show.legend = FALSE, position = 'identity') +
  scale_fill_highcontrast() +
  scale_color_highcontrast() +
  scale_x_continuous(paste0('Temperature \U00B0', 'C'), expand = c(0, 0)) +
  scale_y_continuous('Fitness', breaks = NULL, limits = c(0, 0.16), expand = c(0, 0))
```

During the last century, changes in climate in British Columbia (BC), Canada, have resulted in widespread warming, milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@eberle_technical_2022]. These changes are expected continue over the next 100 years and compound their effects on mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress and human pressure [@deb_modelling_2020; @woo-durand_increasing_2020; @weststrate_how_2024].

This paper provides a BC-focused analysis on the effects of proximate weather and long-term climate on the movement of terrestrial mammal species. Using over two decades of telemetry data and proximal hourly weather data, we estimate how animals altered their movement frequency, movement speed, and habitat selection in response to air temperature. We then pair these responses with climate change projections to forecast changes mammals movement in the current century under different climate change scenarios [@riahi_shared_2017]. The analyses include data from six species (*Canis lupus*, *Cervus elaphus*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; see Table \@ref(tab:data-table)) across a large spatial range of British Columbia (Fig. \@ref(fig:bc-map)). We discuss the consequences of changes in mammalian movement in relationship to energetics, food encounter rates, and the incidences of human-wildlife.

@@

```{r data-table}
readRDS('../models/movement-models-2024-06-06.rds') %>%
  transmute(Dataset = if_else(grepl('Puma', dataset_name),
                                'Puma concolor', dataset_name) %>%
              stringr::str_replace_all('_', ' ') %>%
              stringr::str_replace_all(' boreal', ' (boreal)') %>%
              stringr::str_replace_all(' southern mountain',
                                       ' (southern mountain)'),
            animal,
            tel = purrr::map(tel, data.frame)) %>%
  unnest(tel) %>%
  group_by(animal) %>%
  mutate(dt = timestamp - lag(timestamp)) %>%
  group_by(Dataset) %>%
  summarize(Start = min(as.Date(timestamp)),
            End = max(as.Date(timestamp)),
            dt = as.numeric(median(dt, na.rm = TRUE))) %>%
  mutate(dt = round('hours' %#% dt, 2)) %>%
  rename(`Median sampling interval (hours)` = dt) %>%
  knitr::kable(format = 'pipe', caption = 'Start and end of the GPS telemetries along with the median sampling interval for each of the six species included in this study.') %>%
  kableExtra::column_spec(1, italic = TRUE) %>%
  kableExtra::kable_styling(font_size = 10)
```

@@

Animal telemetry datasets were obtained from multiple sources. Elk tracking data were obtained from Movebank [@kays_movebank_2022; study name: Elk in southwestern Alberta, see @ciuti_human_2012], while the boreal caribou and wolf data were obtained from a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/)). ***All other tracking data were obtained from private collaborators*** [@refs_to_studies]. Outlier GPS locations were removed following diagnostic analyses of the location of each point as well as straight-line displacement and turning angle between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$), especially if points before or after the potential outlier indicated stationary behavior. The script used to clean the data and all custom functions used are available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature)).

@@

Rasters of three-hour reanalyzed air temperature data were downloaded from the ERA5 dataset [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)). Proximal weather conditions were then estimated for each location by extracting the values from the corresponding cell from the temporally nearest raster using the `extract()` function from the `terra` package for `R` [version 1.7-46, @hijmans_terra_2023].

@@

Since `ctmm`'s movement models allow for stochastic speeds but assume constant movement, speeds were set to zero if they were not significantly different from zero ($\alpha = 0.05$), under the assumption of a $\chi$-distributed speeds. ***NOT SURE IF THIS IS CORRECT*** The function used is available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R)). To avoid artifacts due to excessively small, non-zero speeds, estimated speeds were then clustered into two groups using a $k$-means algorithm with two centers (see Appendix A).

@@

We estimated the effects of hourly temperature on mammals' movement using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019] via `mgcv` package for `R`. The first HGAM had a binomial family of distribution and estimated the probability of a mammal moving as a function of time of day, day of year, and temperature. Since there was substantial variation in the direction and wiggliness of each species' trends, the effects were modeled independently `by` smooths for each species [see model I in Figure 4 of @pedersen_hierarchical_2019]. The model accounted for the cyclicity of time of day and day of year using cyclic cubic splines, which provide continuity in the smooth term by ensuring the function and its slope (first derivative) and curvature (second derivative) are continuous between consecutive cycles. The second HGAM had the same structure, but it used a Gamma family of distributions to estimate the speed of an animal given that it was moving. Together, the two HGAMs inform us on an animal's long-term speed (i.e., distance traveled per unit time), since it is the product of probability that the animal is moving and the speed it moves at, given that it is moving.

@@

An animal's habitat selection is proportional to the number of times one can observe it in an environment. Consequently, one can view the spatial distributions of an animal in space as a poisson process whose mean depends on the amount of space, $\Delta u$, and time, $\Delta t$, the observation takes place over, which we can write as $\lambda(\Delta u, \Delta t)$. If we shrink both $\Delta u$ and $\Delta t$ to a single point in space and a single moment in time, we now have $\lambda(u, t) \le 1$, since we cannot observe more than one animal in a given point in time and spce $(u, t)$. Additional theoretical background is provided by @aarts_estimating_2008 and @alston_mitigating_2022 and in Appendix B.

@@

* energetics
* encounter rates (lower encounter rates with food [@hou_cold_2020])
* HWI
